---
title: "cbDVG downstream analysis"
author: "Dr Biruhalem"
date: "2023-12-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## loading R package libraries 

```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(reticulate)
library(ggsignif)
library(gridExtra)
```

#-------------
## Loading data
#-------------

```{r}

dvg <- read.csv("D:/SSPE_raw.fatsq_files/DI-tector/merged_output_V2_all.csv")  # merged_output.txt was saved as merged_output_V2_all.csv and loaded 

head(dvg)

```

#---------------------------
## Sumerizing the DVG counts
#---------------------------

#Sumerizing the counts of forward and reverse strand to total_count. For a given DVG in each sample, if the DVG has a swich in BP_pos and RI_pos in the same sample, then add the Count values of both to total_count. Else keep the Count value as total_count


```{r}
# Create the 'strand' column
dvg <- dvg %>%
  mutate(strand = ifelse(BP_Pos - RI_Pos < 0, "minus", "plus"))

dvg <- dvg %>%
  mutate(
    reverse = sapply(1:nrow(dvg), function(i) {
      row <- dvg[i, ]
      any(dvg$Samples == row$Samples & dvg$BP_Pos == row$RI_Pos & dvg$RI_Pos == row$BP_Pos)
    })
  )

# Create the 'total_count' column within each sample
dvg$total_count <- sapply(1:nrow(dvg), function(i) {
  row <- dvg[i, ]
  sum_counts <- sum(dvg$Counts[dvg$Samples == row$Samples & dvg$BP_Pos == row$RI_Pos & dvg$RI_Pos == row$BP_Pos])
  return(row$Counts + sum_counts)
})

# Sort the data frame by 'total_count' in descending order
dvg <- dvg[order(-dvg$total_count), ]

head(dvg, 100)

table(dvg$strand, dvg$reverse)

```

```{r}
head(dvg, 100)
```

##Filtering unique DVG. For DVG with bith plus and minus strand, select the minus strand. DVGs with only plus or minus strand will be aslo kept as unique DVG.

#selecting data with minus strand if exist

```{r}
dvg1 <- dvg %>%
  filter(ifelse(reverse, strand == "minus", TRUE))
head(dvg1)

table(dvg1$strand, dvg1$reverse)

#save the file

write.csv(dvg1, "C:/Users/user/Desktop/SSPE_DItector/DItector/dvg1.csv")

```

#---------------------------------------------------------------
## Extracting 5'cbDVGs with rule of six and more than 100 copies 
#---------------------------------------------------------------

```{r}
dvg_filtered <- subset(dvg1, Rule_of_six == "YES" & total_count > 100)
dim(dvg_filtered)

dvg_filtered_cb <- subset(dvg_filtered, DVG.stype == "5'cb/sbDVG")

dim(dvg_filtered_cb)

write.csv(dvg_filtered_cb, "C:/Users/user/Desktop/SSPE_DItector/DItector/dvg_filtered_cb.csv")

```

#extracting the 276 cbDVGs for heatmap


```{r}
pivot_df <- dvg_filtered_cb %>%
  pivot_wider(
    id_cols = BP_Pos_RI_Pos,
    names_from = Samples,
    values_from = total_count,
    values_fill = 0  # Replace missing values with 0
  )

# Print the result
head(pivot_df)
```


```{r}
# 
pivot_df <- dvg_filtered_cb %>%
  pivot_wider(
    id_cols = BP_Pos_RI_Pos,
    names_from = Samples,
    values_from = total_count,
    values_fill = 0  # Replace missing values with 0
  ) %>%
  mutate(Length = dvg_filtered_cb$Length[match(BP_Pos_RI_Pos, dvg_filtered_cb$BP_Pos_RI_Pos)])

# Print the result
head(pivot_df)

write.csv(pivot_df, "C:/Users/user/Desktop/SSPE_DItector/DItector/pivot_df.csv")

```

#----------------------------
# Scater plots for Figure 4B-H
#--------------------------------

```{r}
library(ggplot2)
library(RColorBrewer)
library(extrafont)

# Load fonts
loadfonts(device = "win")

# Assuming cb_data is your data frame
pdf("cb_data13.pdf")

# Find the overall minimum and maximum values of total_count across all samples
overall_min_count <- min(cb_data_2$total_count, na.rm = TRUE)
overall_max_count <- max(cb_data_2$total_count, na.rm = TRUE)

# Define the DVG_IDs and their corresponding cbDVG_IDs
highlighted_DVG_IDs <- c("DVG_324_15675_15791", "DVG_336_15652_15802", "DVG_342_15655_15793", "DVG_342_15654_15794", "DVG_426_15570_15794", "DVG_1224_14822_15744", "DVG_1974_14081_15735")
cbDVG_IDs <- c("cbDVG_324(104/116)", "cbDVG_336(93/150)", "cbDVG_342(102/138)", "cbDVG_342(101/140)", "cbDVG_426(101/224)", "cbDVG_1224(151/922)", "cbDVG_1974(160/1654)")

# Create a mapping of DVG_IDs to cbDVG_IDs
DVG_ID_to_cbDVG_ID <- setNames(cbDVG_IDs, highlighted_DVG_IDs)

# Create a color palette for the DVG_IDs
num_highlighted <- length(highlighted_DVG_IDs)
#palette <- brewer.pal(n = num_highlighted, name = "Set1")

palette <- c("orange","red","blue", "lightblue","white","yellow","black")

# Create a mapping of cbDVG_IDs to consistent colors
consistent_colors <- setNames(palette, cbDVG_IDs)

# Function to rename DVG_IDs in plot_data
rename_DVG_IDs <- function(df, mapping) {
  df$DVG_ID <- mapping[df$DVG_ID]
  return(df)
}

for (sample in unique(cb_data_2$Samples)) {
  plot_data <- subset(cb_data_2, Samples == sample)
  
  # Ensure plot_data is not empty
  if (nrow(plot_data) == 0) {
    next
  }
  
  # Rename DVG_IDs in the plot data
  plot_data <- rename_DVG_IDs(plot_data, DVG_ID_to_cbDVG_ID)
  
  # Separate highlighted and non-highlighted data
  highlighted_data <- plot_data[plot_data$DVG_ID %in% cbDVG_IDs, ]
  non_highlighted_data <- plot_data[!plot_data$DVG_ID %in% cbDVG_IDs, ]

  # Use a size gradient for total_count with overall scale
  p_cb_2 <- ggplot() +
    geom_point(data = non_highlighted_data, aes(x = BP_Pos, y = RI_Pos, size = total_count, color = DVG_ID), alpha = 1) +
    geom_point(data = highlighted_data, aes(x = BP_Pos, y = RI_Pos, size = total_count, color = DVG_ID), alpha = 1) +
    labs(title = paste("cbDVG scater plot of BP and RI", sample),
         x = "BP_Pos",
         y = "RI_Pos",
         size = "Total Count",
         color = "cbDVG_ID") +
    coord_fixed(ratio = 1, xlim = c(13800, 15894), ylim = c(13800, 15894)) +  # Ensure equal scaling for x and y axis and set axis limits
    scale_x_continuous(breaks = seq(13894, 15894, by = 500)) +  # Set x-axis ticks
    scale_y_continuous(breaks = seq(13894, 15894, by = 500)) +  # Set y-axis ticks
    scale_size_continuous(range = c(1, 10), limits = c(overall_min_count, overall_max_count), breaks = seq(overall_min_count, overall_max_count, length.out = 6),
                          labels = scales::number_format(scale = 1)) +  # Round up to integers
    scale_color_manual(values = consistent_colors) +  # Assign consistent colors based on cbDVG_ID
    theme(text = element_text(family = "sans", size = 12),
          plot.title = element_text(family = "sans", size = 12, hjust = 0.5),
          axis.title = element_text(family = "sans", size = 12),
          axis.text = element_text(family = "sans", size = 12),
          legend.title = element_text(family = "sans", size = 12),
          legend.text = element_text(family = "sans", size = 12),
          axis.title.y = element_text(margin = margin(r = 10))) +  # Adjust the margin of the Y-axis title
    guides(color = guide_legend(override.aes = list(size = 5)))  # Increase the size of the legend colored circles
  
  # Print the plot to the PDF device
  print(p_cb_2)
}

# Close the PDF device
dev.off()

```





